---
title: "Course Project 1(Reproducible Research)"
output:
  html_document:
    keep_md: true
---

### Question 1. Code for reading in the dataset and/or processing the data.

Change the working directory where data is stored. Unzip the data and load the data in Rstudio.Steps given below - 

```{r, echo=TRUE}
unzip('repdata_data_activity.zip')
activity <- read.csv('activity.csv', colClasses = c('numeric', 'Date', 'numeric'))


```

Initial look of the data - 

```{r, echo=TRUE}
head(activity)
str(activity)
summary(activity)
```

### Question 2. What is the mean total number of steps taken per day?

As mentioned, for this part of assignment, we can ignore the missing values in the dataset.
Calculating the total number of steps taken per day -

```{r, echo=TRUE}
data <- aggregate(steps ~ date, data = activity, FUN = sum)
library(ggplot2)
```

Histogram of the total number of steps taken each day - 

```{r, echo=TRUE}
ggplot(data, aes(x=steps)) + geom_histogram(color='blue', fill="skyblue", bins = 20) + labs(title = 'Histogram of Total Number of Steps taken each day', x = 'Total Steps', y = 'Count')
```

Mean and median of the total number of steps taken per day - 

```{r, echo=TRUE}
library(dplyr)
mean_median <- summarize(data, mean_steps = mean(steps), median_steps = median(steps))
mean_median
```

### Question 3. What is the average daily activity pattern ?



```{r, echo=TRUE}
average_steps <- activity %>% group_by(interval) %>% summarize(mean_steps = mean(steps, na.rm = TRUE))
plot(x = average_steps$interval, y = average_steps$mean_steps, type = 'l', main = 'Time Series Plot of Average Steps taken per Interval', xlab = 'Intervals(5 min)', ylab = 'Number of Steps', col = 'red', lwd = 1.5, family = 'serif')
```

5-minute interval, which contains the maximum number of steps on average across all the days in the dataset - 

```{r, echo=TRUE}
average_step <- arrange(average_steps, desc(mean_steps))
average_step[1, ]
```

The interval with the maximun number of steps is shown in the interval column above.

### Question 4. Imputing missing values.

The total number of missing values in the dataset (i.e. the total number of rows with \color{red}{\verb|NA|}NAs) are -

```{r , echo=TRUE}
sum(rowSums(is.na(activity)))
```

Filling the missing values of the dataset with the mean for that 5 min interval  and creating a new dataset "activity1"- 

```{r, echo=TRUE}
activity1 <- activity
for (i in 1:dim(activity1)[1]){
        if (is.na(activity1[i, 1]) == TRUE ){
                activity1[i, 1] = average_steps[average_steps$interval %in% activity1[i,3], 2]
        }  
}
head(activity1)
```

Histogram of the total number of steps taken each day and mean, median of total number of steps taken per day - 

```{r, echo=TRUE}

total_steps_per_day <- aggregate(steps~date, data = activity1, FUN = sum)
ggplot(total_steps_per_day, aes(x=steps)) + geom_histogram(color='blue', fill="skyblue", bins = 20) + labs(title = 'Histogram of Total Number of Steps taken each day(Imputed Data)', x = 'Total Steps', y = 'Count')
mean_median_1 <- summarize(total_steps_per_day, mean_steps = mean(steps), median_steps = median(steps))
mean_median_1
```

Impact of imputing missing data on the estimates of the total daily number of steps -

```{r, echo=TRUE}
comparison <- rbind(mean_median, mean_median_1)
Dataset <- c('Dataset with Missing Values', 'Dataset with Imputed Values')
comparison <- cbind(comparison, Dataset)
col_order <- c('Dataset', 'mean_steps', 'median_steps')
comparison <- comparison[, col_order]
comparison
```

Also as seen from both the histograms of the dataset with missing values and dataset with imputed values, there is change in the frequency of the number of steps.

```{r,echo=TRUE}
na_grouping <- activity[is.na(activity$steps), c('date', 'steps')]
 table(na_grouping$date)
```
So there were 8 full days with NA values, so the analysis without imputed values were done without considering these 8 days.

### Question 5. Are there differences in activity patterns between weekdays and weekends?

Creating a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day - 

```{r, echo=TRUE}

day_of_week <- weekdays(activity1$date)
day_of_week_new <- c()
for (i in 1:dim(activity1)[1]){
        if (day_of_week[i] %in% c('Saturday', 'Sunday')){
                day_of_week_new[i] = 'Weekend'
        }
        else{
                day_of_week_new[i] = 'Weekday'
        }
}
activity1$day <- day_of_week
activity1$day_in_week <- day_of_week_new
activity1$day_in_week <- as.factor(activity1$day_in_week)
activity1 <- activity1[-c(4)]
str(activity1)
head(activity1)
```

Time series plot of the 5-minute interval and the average number of steps taken, averaged across all weekday days or weekend days-

```{r,echo=TRUE}
data_day_in_week <- activity1 %>% group_by(day_in_week, interval) %>% summarize(mean_steps = mean(steps))
qplot(x = interval, y = mean_steps, data = data_day_in_week, facets = ~day_in_week, geom = c('line'), xlab = '5-min interval', ylab = 'Number of steps')
```
